---
title: "Temporal evolution of forest habitat amount and fragmentation in Europe (1990–2018)"
author: "Cyrille Marchant"

format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show/Hide code"
    theme: cosmo
    self-contained: false

server: shiny

params:

  # ---------------------------------------------------------
  # STUDY AREA
  # ---------------------------------------------------------

  study_regions:
    - "ESP"
    - "EST"
    - "LVA"
    - "LTU"

  years:
    - 1990
    - 2000
    - 2006
    - 2012
    - 2018

  # ---------------------------------------------------------
  # LANDSCAPE ANALYSIS
  # ---------------------------------------------------------

  grid_km: 10

  habitat_focus: "forest"

  forest_classes:
    - 311
    - 312
    - 313

  # ---------------------------------------------------------
  # COMPUTATION
  # ---------------------------------------------------------

  use_cache: true

  # ---------------------------------------------------------
  # GIT / GITHUB
  # ---------------------------------------------------------

  push_to_github: false

  git_commit_message: "Update habitat fragmentation workflow"

execute:
  echo: true
  warning: false
  message: false
---

# Introduction

Landscape fragmentation is a major component of landscape change and can affect habitat availability, ecological connectivity and biodiversity. Changes in land cover can modify both the total amount of habitat and its spatial configuration.

This project investigates the temporal evolution of forest habitat amount and fragmentation between 1990 and 2018 using CORINE Land Cover data.

The analysis compares two European regions with contrasting geographic and landscape characteristics: Spain and the Baltic States (Estonia, Latvia and Lithuania). A regular **10 × 10 km grid** is used as the spatial unit of analysis. The use of equal-sized cells allows landscape metrics to be compared independently of the size and geometry of administrative units.

The complete workflow is designed to be reproducible. Study areas, years, grid resolution and forest classes are defined as parameters at the beginning of the document and can therefore be modified without rewriting the analysis.

## Research question

**How did forest habitat amount and landscape fragmentation evolve between 1990 and 2018 in Spain and the Baltic States, and do these regions show different trajectories of landscape change?**

## Objectives

The analysis has four main objectives:

1.  quantify changes in forest habitat amount between 1990 and 2018;
2.  quantify changes in forest spatial configuration and fragmentation;
3.  identify spatial hotspots of forest loss, gain and fragmentation;
4.  compare temporal trajectories between Spain and the Baltic States.

# Project initialization

## Required packages

The analysis combines vector and raster geoprocessing, landscape ecology metrics, data manipulation and interactive visualization.

`sf` is used for vector spatial data, `terra` for raster processing, and `landscapemetrics` for the computation of landscape metrics. European administrative boundaries are obtained with `giscoR`. The `here` package is used to construct relative file paths, avoiding computer-specific working directories.

```{r}
#| label: setup-packages

# Install pacman only if it is not already installed
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

# Install missing packages and load all required packages
pacman::p_load(
  here,
  fs,

  dplyr,
  tidyr,
  purrr,
  stringr,
  tibble,

  sf,
  terra,

  giscoR,

  landscapemetrics,
  exactextractr,

  ggplot2,
  patchwork,

  leaflet,
  shiny,

  gert
)

# Raster processing options
terra::terraOptions(
  progress = 1
)
```

## Reproducible project root

No absolute working directory is used in this project.

Instead of using `setwd()`, all paths are constructed from the root of the RStudio project with `here::here()`. Consequently, another user can clone the GitHub repository in any directory and run the workflow without modifying local paths.

```{r}
#| label: check-project-root

project_root <- here::here()

cat(
  "Project root detected by here:\n",
  project_root,
  "\n"
)
```

The expected project root on the author's computer is currently:

``` text
C:/Users/Cyril/OneDrive - UCL/Documents/habitat-fragmentation-europe
```

This path is **not hard-coded in the analysis** and is shown only for information.

## Project parameters

All major methodological parameters are retrieved from the YAML header. These values constitute the main settings that can be modified to reproduce the workflow for other areas or spatial resolutions.

```{r}
#| label: analysis-parameters

study_regions  <- unlist(params$study_regions)
years          <- as.integer(unlist(params$years))
grid_km        <- as.numeric(params$grid_km)
habitat_focus  <- params$habitat_focus
forest_classes <- as.integer(unlist(params$forest_classes))
use_cache      <- isTRUE(params$use_cache)

analysis_parameters <- tibble::tibble(
  parameter = c(
    "Study regions",
    "Years",
    "Grid resolution",
    "Habitat focus",
    "CORINE forest classes",
    "Cache"
  ),
  value = c(
    paste(study_regions, collapse = ", "),
    paste(years, collapse = ", "),
    paste0(grid_km, " × ", grid_km, " km"),
    habitat_focus,
    paste(forest_classes, collapse = ", "),
    as.character(use_cache)
  )
)

analysis_parameters
```

## Reproducible project structure

The project directories are automatically created when the document is executed. This prevents the analysis from depending on folders manually created on a specific computer.

```{r}
#| label: project-structure

paths <- list(
  data_input = here::here("data_input"),
  corine     = here::here("data_input", "corine"),
  boundaries = here::here("data_input", "boundaries"),

  data_output = here::here("data_output"),
  figures     = here::here("data_output", "figures"),
  maps        = here::here("data_output", "maps"),
  tables      = here::here("data_output", "tables"),

  cache       = here::here("cache")
)

purrr::walk(
  paths,
  fs::dir_create
)

tibble::tibble(
  folder = names(paths),
  path   = unlist(paths),
  exists = purrr::map_lgl(paths, fs::dir_exists)
)
```

The resulting project structure is:

``` text
habitat-fragmentation-europe/
│
├── habitat-fragmentation-europe.Rproj
├── Workflow final.qmd
├── README.md
├── .gitignore
│
├── data_input/
│   └── corine/
│
├── data_output/
│   ├── figures/
│   ├── maps/
│   └── tables/
│
└── cache/
```

## Check that the folders exist

This diagnostic verifies that the complete project structure has been successfully generated.

```{r}
#| label: check-folders

folder_check <- tibble::tibble(
  folder = names(paths),
  exists = purrr::map_lgl(
    paths,
    fs::dir_exists
  )
)

folder_check
```

# Version control and GitHub

## Git ignore configuration

Raw CORINE Land Cover rasters can be very large and should not normally be stored directly in the GitHub repository.

The `.gitignore` file therefore excludes raw CORINE files and temporary cache files while retaining the Quarto source code and derived lightweight outputs.

```{r}
#| label: create-gitignore

gitignore_file <- here::here(".gitignore")

gitignore_lines <- c(

  "# ----------------------------------",
  "# R / RStudio",
  "# ----------------------------------",
  ".Rhistory",
  ".RData",
  ".Ruserdata",
  ".Rproj.user/",

  "",

  "# ----------------------------------",
  "# Quarto",
  "# ----------------------------------",
  ".quarto/",
  "*_files/",

  "",

  "# ----------------------------------",
  "# Temporary computation cache",
  "# ----------------------------------",
  "cache/",

  "",

  "# ----------------------------------",
  "# Large raw CORINE datasets",
  "# ----------------------------------",
  "data_input/corine/*.tif",
  "data_input/corine/*.tiff",
  "data_input/corine/*.zip",

  "",

  "# ----------------------------------",
  "# Operating system files",
  "# ----------------------------------",
  ".DS_Store",
  "Thumbs.db"
)

writeLines(
  gitignore_lines,
  gitignore_file
)

cat(
  ".gitignore created at:\n",
  gitignore_file
)
```

## Git repository diagnostic

The following code checks whether the current project is already initialized as a Git repository.

```{r}
#| label: git-diagnostic

git_directory <- here::here(".git")

is_git_repository <- dir.exists(git_directory)

if (is_git_repository) {

  cat("Git repository detected.\n\n")

  git_status_current <- gert::git_status(
    repo = here::here()
  )

  print(git_status_current)

} else {

  cat(
    "The project directory is not yet initialized as a Git repository.\n"
  )
}
```

## Git identity

The following diagnostic displays the Git identity currently configured on the computer.

```{r}
#| label: git-identity

git_config <- gert::git_config_global()

git_identity <- git_config |>
  dplyr::filter(
    name %in% c("user.name", "user.email")
  ) |>
  dplyr::select(
    setting = name,
    value
  )

if (nrow(git_identity) == 2) {

  cat("Git identity is correctly configured.\n")
  print(git_identity)

} else {

  cat(
    "Git identity is not completely configured.\n",
    "Configure user.name and user.email before committing.\n"
  )
}
```

# Data

## Study area

The study area consists of Spain and the three Baltic States:

-   Spain (`ESP`);
-   Estonia (`EST`);
-   Latvia (`LVA`);
-   Lithuania (`LTU`).

European boundaries will be downloaded directly from GISCO rather than manually stored in the project. This improves reproducibility and makes it possible to adapt the workflow to other European countries.

*The next section constructs the study area and the common 10 × 10 km analytical grid.*
